Report ASP20 Boost

### 2020-06-04 Reports/report_1/1_REPORT.Rmd
update report_1

@@

In part 2 of this report we explain the concept of boosting and elaborate on how it suites to address the problem of location-scale regression. 
In the following part 3 we describe our implementation milestones, whereby part 4 especially elaborates on componentwise boosting. 
In part 5 we present our thoughts on upcoming challenges and critical aspects of the package development.

@@

The first step in our progress was to understand the theoretical concept of boosting, and which types of problems it is able to solve. In the following the term "booster" refers to a boosting algorithm.

@@

To explain the concept and follow our process of understanding we first explain boosting for location parameters. Boosting scale parameters requires another way of thinking about it and constituted a first milestone in our understanding and implementation.
The boosting concept relies on the idea of iterative estimation. 
To estimate the location parameters beta via boosting, one starts with initial estimates, which may be far from optimal. 
However, with estimates and the response vector at hand, residuals can be calculated. 
The boosting-algortihm focuses on those calculated residuals.
The effect of the predictors xi on these residuals is estimated by least squares and the resulting effect is added to the location-estimate beta, adjusted by some learning rate. 
This yields a better fit of the location-model, which results in smaller residuals.
These new residuals are estimated again in the next iteration, yielding smaller effect sizes, and hence a convergence of the boosting algorithm towards the OLS-estimate.
A simple booster stops after a fixed number of iterations is reached.
**(comp. Kneib et al)**

@@

[Loss-function erklären: Boosting can be viewed as lowering iteratively the amount of a loss function, which is in our case the score of the normal distribution]

The concept for boosting scale parameters is the same. However, residuals for the variance estimates can not be calculated, since there is no such thing as "observed" variances for individual response values. The solution is to substitute the calculated residuals with a different term: the derivative of the loss function, which in our case is the log-likelihood of the normal distribution. Instead of estiamting stepwise residuals, one estiamtes hence stepwise the score.

The resulting loss fcuntion then, is defined by sum of squares oa the term:
ui (=deviance-residuals) - inverse predicted scores * working-residuals

We employed the resid-function of the asp20model class, on which our class is built on.
<!-- The working and response residuals are the observed values -->
<!-- minus the fitted values for the location. -->
<!-- The deviance and Pearson residuals are weighted residuals, -->
<!-- divided by the fitted values for the scale. -->

@@

In the current version of the asp20boost-package, the boosting algorithm is already implemented in order to allow componentwise boosting, which will be explained in part 4. 
For good explanation and for following our development process in this part we first describe the implementation of ordinary boosting - which however does not appear in our code anymore.

A simple booster for the location parameters beta works without extending the LocationScaleRegression-Class. 
The current residuals may be extrated with the resid()-method, then estimated. The location parameter is updated. 
This causes the LocationScaleRegression-Class to calculate updated state-dependent objects, such as the loglikelihood, gradients and the residuals. 
This is what's needed to repeat the procedure of estimating residuals and updating location parameters.

The simple booster for gamma consists in the same two repeating steps of first estimating a term, and second updating the scale parameters gamma by adding the estimated effect - adjusted by a learning rate. 
The term to be estimated, however, is not the residual, but the derivative of the loglikelihood of the normal distribution. 
To calculate this, our code makes use of the resid-function(), but this time passing the argument "deviance", which results in residuals adjusted by the fitted scale estimates.

[Is the gamma booster dependent on simultaneous running of beta booster?]

@@

A usefull functionality of boosting is componentwise boosting. 
The idea is to not update a whole parameter vector, but only one entry of it.
The entry chosen for the update at this juncture is the one yielding the best improvement in the sense of lowering the loss function.
The loss function in our case is the loglikelihood of the normally distributed data.
Hence in each iteration only one component of the location-, as well as one component of the scale-parameters are estimated.

The implementation in our package works via the extension of the LocationScaleRegressionClass by two active fields "bestFittingVariableBeta" and "bestFittingVariableGamma". 
These functions partition the design matrix X, respectively Z, into its single columns and then estimate the residuals - respectively the score - seperately for each component, and determine loss functions for a hypothetical update with the respective component. 
Comparing with the old loss function value, the highest loss-improvement may be determined and the respective component is used to update the parameter-vector.

@@

The most important functionality we intend to implement in our package is variable selection via cross validation. Working out the theoretical concept behind this idea is one of our next milestones.

@@

* A major performance problem in our code is that small learning rates for the beta-booster cause processing time of the code to increase sharply. This remains to be solved.

*A good choice of starting values for beta and gamma may enhance performance. A possible candidate is the mean of the response.

@@

* Document the extension of the R6-Class properly and inherit documentation of the LocationScaleRegression-Class.

>>>>

**[S: please format all equations and terms - y, xi with a subscripted i, zi (same), ...]**

@@

In *part 2* of this report we explain the concept of boosting and elaborate on how it suites to address the problem of location-scale regression. 
In the following *part 3* we describe our implementation milestones, whereby *part 4* especially elaborates on componentwise boosting. 
In *part 5* we present our thoughts on upcoming challenges and critical aspects of the package development.

@@

**[S: please format all terms - xi, beta and gamma as greek letters, v also as greek letter ...]**
**[S: please add reference]**

The first step in our progress was to understand the theoretical concept of boosting, and which types of problems it is able to solve. To explain the concept and follow our process of understanding we first explain boosting for location parameters. 
Boosting scale parameters requires another way of thinking about it and constituted a first milestone in our understanding and implementation.
In the following, the term "booster" refers to a boosting algorithm.

@@

The boosting concept relies on the idea of iterative estimation. 
To estimate the location parameters beta via boosting, one starts with initial estimates, which may be far from optimal. 
However, with estimates and the response vector at hand, residuals can be calculated. 
The boosting algortihm focuses on those calculated residuals.
The effect of the predictors xi on these residuals is estimated by least squares and the resulting effect is added to the location-estimate beta, adjusted by some learning rate *v*. 
This yields a better fit of the location-model, which results in smaller residuals.
These new residuals are estimated again in the next iteration, yielding smaller effect sizes, and hence a convergence of the boosting algorithm towards the OLS-estimate.
A simple booster stops after a fixed number of iterations is reached.
**(compare Kneib et al, p217,218)**

@@

A boosting algorithm for scale parameters **gamma** consists in the same two repeating steps of first estimating a term, and second updating **gamma** by adding the estimated effect - adjusted by a learning rate. 
The term in the estimating step is, however, not the residual.
For boosting **gamma** this term equals the score function of the model, that is the derivative of the loglikelihood of the normal distribution.

**[Why is that? kurze Erklärung]**

<!-- The concept for boosting scale parameters is the same. However, residuals for the variance estimates can not be calculated, since there is no such thing as "observed" variances for individual response values. The solution is to substitute the calculated residuals with a different term: the derivative of the loss function, which in our case is the log-likelihood of the normal distribution. Instead of estiamting stepwise residuals, one estiamtes hence stepwise the score. -->

<!-- The resulting loss fcuntion then, is defined by sum of squares oa the term: -->
<!-- ui (=deviance-residuals) - inverse predicted scores * working-residuals -->

<!-- We employed the resid-function of the asp20model class, on which our class is built on. -->
<!-- The working and response residuals are the observed values -->
<!-- minus the fitted values for the location. -->
<!-- The deviance and Pearson residuals are weighted residuals, -->
<!-- divided by the fitted values for the scale. -->

@@

The implementation of the boosting algorithm is accomplished by a function *gradient_boost*, which harmonizes with the R6-Class *LocationScaleRegressionBoost*. 
This class, in turn, is built upon the *LocationScaleRegression* class of the `asp20model`-package. 
It follows a strict dependency of the `asp20boost`- on the `asp20model`-package.

In the current version of the `asp20boost`-package, the boosting algorithm is already implemented in order to allow componentwise boosting, which will be explained in part 4. 
For good explanation and for following our development process in this part we first describe the implementation of ordinary boosting - which however does not appear in our code anymore.

A simple booster for the location parameters **beta** works without extending the *LocationScaleRegression* class. 
The current residuals may be extrated with the *resid()*-method, then estimated. The location parameter is updated. 
This causes the *LocationScaleRegression* class to calculate updated state-dependent objects, such as the loglikelihood, gradients and the residuals. 
This is what's needed to repeat the procedure of estimating residuals and updating location parameters **Beta**.

The simple booster for the scale parameters **gamma**, as already mentioned, consists in the same two repeating steps of first estimating a term, and second updating the scale parameters gamma by adding the estimated effect - adjusted by a learning rate **v**. 
The estimated term here is the derivative of the loglikelihood of the normal distribution.
To calculate this, our code makes use of the resid-function(), but this time passing the argument "deviance", which results in residuals adjusted by the fitted scale estimates.

**[Erklärung weiter ausführen]**

@@

A usefull functionality of boosting is componentwise boosting. 
The idea is to not update a whole parameter vector, but only one entry of it.
The entry chosen for the update at this juncture is the one yielding the best improvement in the sense of lowering the loss function.
The loss function in our case is the loglikelihood of the normally distributed data.
Hence in each iteration only one component of the location-, as well as one component of the scale-parameters are updated.

The implementation in our package works via the extension of the **LocationScaleRegression** class by two active fields *"bestFittingVariableBeta"* and *"bestFittingVariableGamma"*. 
These functions partition the design matrix *X* - respectively *Z* - into its single columns and then estimate the residuals - respectively the scores - seperately for each component, and determine loss functions for a hypothetical update with the respective component. 
Comparing with the old loss function value, the highest loss-improvement may be determined and the respective component is used to update the parameter-vector.

@@

The most important functionality we intend to implement in our package is optimizing the stopping point of our booster. This optimization of the number of iterations may be done via cross validation. Working out the theoretical concept behind this idea is one of our next milestones.

@@

* A major performance problem in our code is that small learning rates for the beta-booster cause processing time of the code to increase sharply. This remains to be solved.
*A good choice of starting values for beta and gamma may enhance performance. A possible candidate is the mean of the response.

@@

* Document the extension of the R6-Class properly and inherit documentation of the LocationScaleRegression-Class applying roxygen2.

### 2020-06-04 Reports/report_1/1_REPORT.tex
update report_1

@@

In part 2 of this report we explain the concept of boosting and
elaborate on how it suites to address the problem of location-scale
regression. In the following part 3 we describe our implementation
milestones, whereby part 4 especially elaborates on componentwise
boosting. In part 5 we present our thoughts on upcoming challenges and
critical aspects of the package development.

@@

The first step in our progress was to understand the theoretical concept
of boosting, and which types of problems it is able to solve. In the
following the term ``booster'' refers to a boosting algorithm.

@@

To explain the concept and follow our process of understanding we first
explain boosting for location parameters. Boosting scale parameters
requires another way of thinking about it and constituted a first
milestone in our understanding and implementation. The boosting concept
relies on the idea of iterative estimation. To estimate the location
parameters beta via boosting, one starts with initial estimates, which
may be far from optimal. However, with estimates and the response vector
at hand, residuals can be calculated. The boosting-algortihm focuses on
those calculated residuals. The effect of the predictors xi on these
residuals is estimated by least squares and the resulting effect is
added to the location-estimate beta, adjusted by some learning rate.
This yields a better fit of the location-model, which results in smaller
residuals. These new residuals are estimated again in the next
iteration, yielding smaller effect sizes, and hence a convergence of the
boosting algorithm towards the OLS-estimate. A simple booster stops
after a fixed number of iterations is reached. \textbf{(comp. Kneib et
al)}

@@

{[}Loss-function erklären: Boosting can be viewed as lowering
iteratively the amount of a loss function, which is in our case the
score of the normal distribution{]}

The concept for boosting scale parameters is the same. However,
residuals for the variance estimates can not be calculated, since there
is no such thing as ``observed'' variances for individual response
values. The solution is to substitute the calculated residuals with a
different term: the derivative of the loss function, which in our case
is the log-likelihood of the normal distribution. Instead of estiamting
stepwise residuals, one estiamtes hence stepwise the score.

The resulting loss fcuntion then, is defined by sum of squares oa the
term: ui (=deviance-residuals) - inverse predicted scores *
working-residuals

We employed the resid-function of the asp20model class, on which our
class is built on.

@@

In the current version of the asp20boost-package, the boosting algorithm
is already implemented in order to allow componentwise boosting, which
will be explained in part 4. For good explanation and for following our
development process in this part we first describe the implementation of
ordinary boosting - which however does not appear in our code anymore.

A simple booster for the location parameters beta works without
extending the LocationScaleRegression-Class. The current residuals may
be extrated with the resid()-method, then estimated. The location
parameter is updated. This causes the LocationScaleRegression-Class to
calculate updated state-dependent objects, such as the loglikelihood,
gradients and the residuals. This is what's needed to repeat the
procedure of estimating residuals and updating location parameters.

The simple booster for gamma consists in the same two repeating steps of
first estimating a term, and second updating the scale parameters gamma
by adding the estimated effect - adjusted by a learning rate. The term
to be estimated, however, is not the residual, but the derivative of the
loglikelihood of the normal distribution. To calculate this, our code
makes use of the resid-function(), but this time passing the argument
``deviance'', which results in residuals adjusted by the fitted scale
estimates.

{[}Is the gamma booster dependent on simultaneous running of beta
booster?{]}

@@

A usefull functionality of boosting is componentwise boosting. The idea
is to not update a whole parameter vector, but only one entry of it. The
entry chosen for the update at this juncture is the one yielding the
best improvement in the sense of lowering the loss function. The loss
function in our case is the loglikelihood of the normally distributed
data. Hence in each iteration only one component of the location-, as
well as one component of the scale-parameters are estimated.

The implementation in our package works via the extension of the
LocationScaleRegressionClass by two active fields
``bestFittingVariableBeta'' and ``bestFittingVariableGamma''. These
functions partition the design matrix X, respectively Z, into its single
columns and then estimate the residuals - respectively the score -
seperately for each component, and determine loss functions for a
hypothetical update with the respective component. Comparing with the
old loss function value, the highest loss-improvement may be determined
and the respective component is used to update the parameter-vector.

@@

\begin{itemize}
\tightlist
\item
  A major performance problem in our code is that small learning rates
  for the beta-booster cause processing time of the code to increase
  sharply. This remains to be solved.
\end{itemize}

*A good choice of starting values for beta and gamma may enhance
performance. A possible candidate is the mean of the response.

@@

\begin{itemize}
\item
  Allow user-input of the LocationScaleRegressionBoost-model in form of
  a dataframe, for example by allowing to pass an optional
  ``data''-argument as known from the lm-call. This may reduce proneness
  to input-errors.
\item
  Visualize the results of the boosted estimates
\item
  Document the extension of the R6-Class properly and inherit
  documentation of the LocationScaleRegression-Class.
\end{itemize}

>>>>

\textbf{{[}S: please format all equations and terms - y, xi with a
subscripted i, zi (same), \ldots{]}}

@@

In \emph{part 2} of this report we explain the concept of boosting and
elaborate on how it suites to address the problem of location-scale
regression. In the following \emph{part 3} we describe our
implementation milestones, whereby \emph{part 4} especially elaborates
on componentwise boosting. In \emph{part 5} we present our thoughts on
upcoming challenges and critical aspects of the package development.

@@

\textbf{{[}S: please format all terms - xi, beta and gamma as greek
letters, v also as greek letter \ldots{]}} \textbf{{[}S: please add
reference{]}}

The first step in our progress was to understand the theoretical concept
of boosting, and which types of problems it is able to solve. To explain
the concept and follow our process of understanding we first explain
boosting for location parameters. Boosting scale parameters requires
another way of thinking about it and constituted a first milestone in
our understanding and implementation. In the following, the term
``booster'' refers to a boosting algorithm.

@@

The boosting concept relies on the idea of iterative estimation. To
estimate the location parameters beta via boosting, one starts with
initial estimates, which may be far from optimal. However, with
estimates and the response vector at hand, residuals can be calculated.
The boosting algortihm focuses on those calculated residuals. The effect
of the predictors xi on these residuals is estimated by least squares
and the resulting effect is added to the location-estimate beta,
adjusted by some learning rate \emph{v}. This yields a better fit of the
location-model, which results in smaller residuals. These new residuals
are estimated again in the next iteration, yielding smaller effect
sizes, and hence a convergence of the boosting algorithm towards the
OLS-estimate. A simple booster stops after a fixed number of iterations
is reached. \textbf{(compare Kneib et al, p217,218)}

@@

A boosting algorithm for scale parameters \textbf{gamma} consists in the
same two repeating steps of first estimating a term, and second updating
\textbf{gamma} by adding the estimated effect - adjusted by a learning
rate. The term in the estimating step is, however, not the residual. For
boosting \textbf{gamma} this term equals the score function of the
model, that is the derivative of the loglikelihood of the normal
distribution.

\textbf{{[}Why is that? kurze Erklärung{]}}

@@

The implementation of the boosting algorithm is accomplished by a
function \emph{gradient\_boost}, which harmonizes with the R6-Class
\emph{LocationScaleRegressionBoost}. This class, in turn, is built upon
the \emph{LocationScaleRegression} class of the
\texttt{asp20model}-package. It follows a strict dependency of the
\texttt{asp20boost}- on the \texttt{asp20model}-package.

In the current version of the \texttt{asp20boost}-package, the boosting
algorithm is already implemented in order to allow componentwise
boosting, which will be explained in part 4. For good explanation and
for following our development process in this part we first describe the
implementation of ordinary boosting - which however does not appear in
our code anymore.

A simple booster for the location parameters \textbf{beta} works without
extending the \emph{LocationScaleRegression} class. The current
residuals may be extrated with the \emph{resid()}-method, then
estimated. The location parameter is updated. This causes the
\emph{LocationScaleRegression} class to calculate updated
state-dependent objects, such as the loglikelihood, gradients and the
residuals. This is what's needed to repeat the procedure of estimating
residuals and updating location parameters \textbf{Beta}.

The simple booster for the scale parameters \textbf{gamma}, as already
mentioned, consists in the same two repeating steps of first estimating
a term, and second updating the scale parameters gamma by adding the
estimated effect - adjusted by a learning rate \textbf{v}. The estimated
term here is the derivative of the loglikelihood of the normal
distribution. To calculate this, our code makes use of the
resid-function(), but this time passing the argument ``deviance'', which
results in residuals adjusted by the fitted scale estimates.

\textbf{{[}Erklärung weiter ausführen{]}}

@@

A usefull functionality of boosting is componentwise boosting. The idea
is to not update a whole parameter vector, but only one entry of it. The
entry chosen for the update at this juncture is the one yielding the
best improvement in the sense of lowering the loss function. The loss
function in our case is the loglikelihood of the normally distributed
data. Hence in each iteration only one component of the location-, as
well as one component of the scale-parameters are updated.

The implementation in our package works via the extension of the
\textbf{LocationScaleRegression} class by two active fields
\emph{``bestFittingVariableBeta''} and
\emph{``bestFittingVariableGamma''}. These functions partition the
design matrix \emph{X} - respectively \emph{Z} - into its single columns
and then estimate the residuals - respectively the scores - seperately
for each component, and determine loss functions for a hypothetical
update with the respective component. Comparing with the old loss
function value, the highest loss-improvement may be determined and the
respective component is used to update the parameter-vector.

@@

\begin{itemize}
\tightlist
\item
  A major performance problem in our code is that small learning rates
  for the beta-booster cause processing time of the code to increase
  sharply. This remains to be solved. *A good choice of starting values
  for beta and gamma may enhance performance. A possible candidate is
  the mean of the response.
\end{itemize}

@@

\begin{itemize}
\item
  Allow user-input of the LocationScaleRegressionBoost-model in form of
  a dataframe, for example by allowing to pass an optional
  ``data''-argument as known from the lm-call. This may reduce proneness
  to input-errors.
\item
  Visualize the results of the boosted estimates
\item
  Document the extension of the R6-Class properly and inherit
  documentation of the LocationScaleRegression-Class applying roxygen2.
\end{itemize}
